An Effective Model of HL-60 Differentiation

### 2017-05-12 Paper_v2.tex
Changes

Sustained mitogen-activated protein kinase (MAPK) activation is a defining feature of ATRA-induced HL-60 differentiation.
ATRA drives sustained activation of the Raf/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
Betraying a feedback loop, MEK inhibition results in the loss of ERK and Raf phosphorylation, and the failure to arrest and differentiate in response to ATRA \cite{Hong2001}.
Retinoic acid (and its metabolites) are ligands for the hormone activated nuclear transcription factors retinoic acid receptor (RAR) and retinoid X receptor (RXR) \cite{Mangelsdorf1990}.
RAR/RXR activation is necessary for ATRA-induced Raf phosphorylation  \cite{Hong2001},
and the formation of an ATRA-inducible signalsome complex at the membrane, which drives MAPK activation.
While the makeup of the signalsome complex is not yet known,
we do know that it is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, plus transcription factors
AhR and IRF1 \cite{Congleton:2012fk,Shen:2011aa,Shen2009,Yen2006,Marchisio:1998aa}.
Signalsome activity is driven by ATRA-induced expression of CD38 and putatively the heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
BLR1 (also known as CXCR5), identified as an early ATRA (or D3)-inducible gene using differential display \cite{YEN1990},
is necessary for MAPK activation and differentiation \cite{WANG2004}, and drives signalsome activity.
Studies of the BLR1 promoter identified a non-canonical RARE site consisting of a 17 bp GT box approximately
1 kb upstream of the transcriptional start that conferred ATRA responsiveness \cite{WANG2004}.
Members of the BLR1 transcriptional activator complex, e.g. NFATc3 and CREB,
are phosphorylated by ERK, JNK or p38 MAPK family members suggesting positive feedback between the signalsome and MAPK activation \cite{Yang2002}.
BLR1 overexpression enhanced Raf phosphorylation and accelerated terminal differentiation, while Raf inhibition reduced BLR1 expression and ATRA-induced terminal differentiation \cite{Wang2008}.
In particular, Raf phosphorylation of the NFATc3 transcription factors at the BLR1 promoter enables transcriptional activation at the RARE by ATRA bound to RAR/RXR \cite{Geil:2014aa}.
BLR1 knock-out cells failed to activate Raf or differentiate in the presence of ATRA \cite{Wang2008}.
Interestingly, both the knockdown or inhibition of Raf, also reduced BLR1 expression and functional differentiation \cite{Wang2008}.
Thus, the expression of signalsome components e.g., BLR1 was Raf dependent, while Raf activation depended upon the signalsome.
A previous computational study of ATRA-induced differentiation of HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient
to explain ATRA-induced sustained MAPK activation, and the expression of a limited number of functional differentiation markers \cite{Tasseff2011}.
Model analysis also suggested that Raf was the most distinct of the MAPK proteins.
However, this previous study developed and analyzed a complex model,
thus leaving open the critical question of what is the minimal positive feedback circuit required to drive ATRA-induced differentiation.

@@

\section*{Results}
We constructed an effective model of ATRA-induced HL-60 differentiation which described signaling and gene expression events following the addition of ATRA (Fig. \ref{fig:network}).
The model connectivity was developed from literature and the studies presented here (Table \ref{tbl:TF-network-connectivity}).
We decomposed the ATRA program into three modules;
a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RAR/RXR (Trigger) signals (Fig. \ref{fig:network}A);
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals (Fig. \ref{fig:network}B); and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors (Fig. \ref{fig:network}C).
Each component of these modules was described by a mRNA and protein balance equation.
Additionally, the signal initiation module also described the abundance of activated species e.g., Trigger and cRaf-pS621 whose values were derived from
unactivated Trigger and cRaf protein levels.
Lastly, because the population of HL-60 cells was dividing, we also considered a dilution term in all balance equations.
The signal initiation module contained nine differential equations, while the signal integration and phenotype modules were collectively encoded by 54 differential equations.
Model parameters were taken from literature (Table \ref{tbl:model-parameters}), or estimated from experimental data using heuristic optimization (see materials and methods).

@@

The signal integration and phenotype modules described ATRA-induced gene expression in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
which then in turn activated the phenotypic program. In particular, Trigger (a surrogate for the RAR$\alpha$/RXR transcriptional complex)
regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and Egr-1.
In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or p47Phox.
We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
We estimated the parameters for the signal initiation, and phenotype modules
from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA \cite{Jensen:2013aa,Jensen:2014aa,Jensen:2015ab,Jensen:2015aa}.
However, the bulk of the model parameters were taken from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
Lastly, we used the \textit{predicted} values of the p21 and E2F protein abundance to estimate a blackbox model of ATRA-induced G0 arrest (Fig. \ref{fig:model-cell-cycle-simulations}).
The phenotype module predicted p21 expression significantly increased and E2F expression decreased, in response to ATRA exposure (Fig. \ref{fig:model-cell-cycle-simulations}A).
We then used the ratio of these values in a polynomial model to calculate the fraction of HL-60 cells in G0 arrest following the addition of ATRA (Fig. \ref{fig:model-cell-cycle-simulations}B).
The third-order polynomial model captured the trend in measured G0-arrest values as a function of time, and was robust to uncertainty in the measured data (Fig. \ref{fig:model-cell-cycle-simulations}B, gray).
Taken together, the output of the signal integration and phenotypic modules was consistent with time-series and steady-state measurements, thereby validating the assumed molecular connectivity.
Moreover, outputs from the phenotype module described the trend in ATRA-induced G0 cell cycle arrest.
Next, we explored which nodes and interactions between nodes in the signal integration module most influenced the system response.

@@

The Gfi-1 and PPAR$\gamma$ proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
We conducted pairwise gene knockout simulations in the signal integration and phenotype modules to estimate which nodes controlled the processing of the Trigger and cRaf-S621 signals.
The difference between the system state with and without the gene knockouts (encoded as a normalized state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
A panel of ten parameter sets was sampled, and the average normalized displacement matrix was decomposed.
The first six modes (approximately 36\% of the total) described $>$95\% of the gene knockout variance,
with the most important components of these modes being the Gfi-1 and PPAR$\gamma$ proteins, and to a lesser extent PU.1, C/EBP$\alpha$ and and AP1 (Fig. \ref{fig:model-response-ko}A).
To better understand which protein-DNA connections were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
Singular value decomposition of the normalized state displacement matrix assembled from the pairwise connection deletions, suggested the first six modes (approximately 26\% of the total) accounted for $>$90\% of the variance.
Globally, the most sensitive interactions controlled p21 and p47Phox expression, markers for cell-cycle arrest and reactive oxygen formation phenotypic axes activated following ATRA addition (Fig. \ref{fig:model-response-ko}B).
Analysis of the modes suggested the action of PPAR$\gamma$, Gfi-1 and C/EBP$\alpha$ were consistently important over multiple target genes.
The connection knockout analysis also revealed robustness in the network.
For example, no pair of deletions qualitatively changed the expression of regulators such as PU.1, Oct1, Oct4 or PPAR$\gamma$.
Thus, the expression of these species was robust to disturbance in the connectivity.
To better understand the combined influence of the PPAR$\gamma$ and Gfi-1 deletions, we computed the fold change in the protein levels in the
single (Gfi-1$^{-/-}$ or PPAR$\gamma$$^{-/-}$) and double (Gfi-1$^{-/-}$ and PPAR$\gamma$$^{-/-}$) mutants for the best fit parameter set (Fig. \ref{fig:model-response-ko-simulations}).
Deletion of Gfi-1 led to a 2-4 fold increase in EGR-1, CD11b and C/EBP$\alpha$ expression, and a $>$8 fold increase in PU.1 abundance (Fig. \ref{fig:model-response-ko-simulations},blue).
On the other hand, deletion of PPAR$\gamma$ led to $>$8 fold downregulation of CD38, p21, IRF1 and Oct1 (Fig. \ref{fig:model-response-ko-simulations},red).
Both knockouts slightly increased E2F expression, but neither influenced the expression of p47Phox.
The double mutant was qualitatively similar to the combined behavior of the two single mutant cases.
Taken together, Gfi-1 and PPAR$\gamma$ controlled the cell-cycle arrest and receptor signaling axes, with PPAR$\gamma$ regulating CD38, IRF1 and p21 expression
while Gfi-1 controlled CD11b expression. These simulations suggested deletion of PPAR$\gamma$ and Gfi-1 would not interfere with reactive oxygen formation,
but would limit the ability of HL-60 cells to arrest.
However, this analysis did not give insight into which components upstream of the signal initiation module were important.
Toward this question, we explored the composition and regulation of the signalsome complex by experimentally interrogating a panel of possible Raf interaction partners.

@@

The composition of the signalsome, and the kinase ultimately responsible for mediating ATRA-induced Raf activation is currently unknown.
To explore this question, we conducted immunoprecipitation and subsequent Western blotting to identify physical interactions between Raf and 19 putative interaction partners.
A panel of 19 possible Raf interaction partners (kinases, GTPases, scaffolding proteins etc)
was constructed based upon known signaling pathways.
We did not consider the most likely binding partner,
the small GTPase RAS, as previous studies have ruled it out in MAPK activation in HL-60 cells \cite{Wang2008,Katagiri1994}.
Total Raf was used as a bait protein for the immunoprecipitation studies.
Interrogation of the Raf interactome suggested Vav1 was involved with ATRA-induced initiation of MAPK activity (Fig. \ref{fig:ip-data}).
Western blot analysis using total Raf and Raf-pS621 specific antibodies confirmed the
presence of the bait protein, total and phosphorylated forms, in the immunoprecipitate (Fig. \ref{fig:ip-data}A).
Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 precipitated with Raf, suggesting a direct physical interaction was possible.
However, only the associations between Raf and Vav1, and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
Furthermore, the Vav1 and Src associations were correlated with Raf-pS621 abundance in the precipitate.
Others proteins e.g., CK2, Akt and 14-3-3, generally bound Raf regardless of phosphorylation status or ATRA treatment.
The remaining 14 proteins were expressed in whole cell lysate (Fig. \ref{fig:ip-data}B),
but were not detectable in the precipitate of Raf IP.
Treatment with the Raf kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 with Raf and Src with Raf (Fig. \ref{fig:ip-data}),
although the signal intensity for Src was notably weak.
However, GW5074 did not influence the association of CK2 or 14-3-3 with Raf, further demonstrating their independence from Raf phosphorylation.
Interestingly, the Raf-Akt interaction qualitatively increased following treatment with GW5074;
however, it remained unaffected by treatment with ATRA.
Src family kinases are known to be important in myeloid differentiation \cite{Miranda2007} and their role in HL-60 differentiation has been investigated elsewhere \cite{Congleton:2012fk}.
Given the existing work and variable reproducibility in the context of the Raf immunoprecipitate,
we did not investigate the role of Src further in this study.
Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with Raf activation following ATRA-treatment.
Previous studies demonstrated that a Vav1-Slp76-Cbl-CD38 complex plays an important role in ATRA-induced MAPK activation and differentiation of HL-60 cells \cite{Shen2009}.
Here we did not observe direct interaction of Raf with Cbl or Slp76;
however, this complex could be involved upstream. Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.

@@

% The presence of Vav1 in Raf/Grb2 complexes has been shown to correlate with increased Raf activity in mast cells \cite{Song1996}.
% Furthermore, studies on Vav1 knockout mice demonstrated that the loss of Vav1 resulted in deficiencies
% of ERK signaling for both T-cells as well as neutrophils \cite{Costello1999,Graham2007}.
% While its function in the signalsome is unclear, Vav1 has been shown to associate with a Cbl-Slp76-CD38 complex in an ATRA-dependent manner;
% furthermore, transfection of HL-60 cells with Cbl mutants that fail to bind CD38, yet still bind Slp76 and Vav1, prevented
% ATRA-induced MAPK activation \cite{Shen2009}.
% Thus, interaction of Cbl-Slp76-Vav1 and CD38 appears to be required for transmission of the ATRA signal by the signalsome.

\section*{Discussion}
In this study, we presented an effective model of ATRA-inducible differentiation of HL-60 cells.
The model consisted of three modules: a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RAR/RXR (Trigger) signals;
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors.
The model described the transcription and translation of genes in each module, and signaling events in each module in a growing population of HL-60 cells.
Model parameters were taken from literature, however, unknown coefficients
that appear in the promoter logic models were estimated from protein measurements in HL-60 cells following ATRA exposure.
Despite its simplicity, the effective model captured key features of the ATRA induced differentiation such as
sustained MAPK activation, and bistability with respect to ATRA exposure.
The model also described the expression of upstream transcription factors which regulated the expression of differentiation markers.
Lastly, analysis of the response of the model to perturbations identified Gfi-1 and PPAR$\gamma$ as master regulators of ATRA-induced differentiation.
We also reported a new ATRA-inducible component of the signalsome, Vav1.
Vav1 is a guanine nucleotide exchange factor for Rho family GTPases that activate pathways leading to actin cytoskeletal rearrangements and transcriptional alterations \cite{Hornstein:2004aa}.
The Vav1/Raf association correlated with Raf activity, was ATRA-inducible and decreased after treatment with the Raf inhibitor GW5074.

Naturally occurring cell fate decisions often incorporate reinforcing feedback and bistability \cite{Ferrell:2002aa,Laslo2006}.
One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
This cascade is activated when oocytes are induced by the steroid hormone progesterone \cite{Xiong:2003aa}.
The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
This is similar to the signal initiation module presented here; ATRA drives signalsome formation, which activates MAPK, which in turn leads to more signalsome activation.
Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, their cell fate programs share a similar architectural feature.
Reinforcing feedback and bistability has also been implicated in hematopoietic cell fate determination.
Laslo et al showed in nonmalignant myelomonocytic cells that the counter antagonistic repressors, Gfi-1 and Egr-1/2 (whose expression is tuned by PU.1 and C/EBP$\alpha$), encode a bistable switch that
results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBP$\alpha$ expression \cite{Laslo2006}.
The current model contained the Gfi-1 and Egr-1/2 agonistic switch; however, its significance was unclear for HL-60 cells.
The expression of Gfi-1, Egr-1/2, C/EBP$\alpha$ and PU.1 was not consistent with the canonical lineage pattern expected from literature.
For example, Egr-1/2 expression (associated with a macrophage lineage) increased, while Gfi-1 expression (associated with a neutrophil lineage)
remained constant following ATRA exposure.
Thus, HL-60 cells, which are a less mature cancer cell line, exhibited a non-canonical expression pattern.
Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
Thus, while reinforcing feedback is often undesirable in human engineered systems, it is at the core of a diverse variety of
cell fate programs and other networks important to human health.

>>>>

Sustained mitogen-activated protein kinase (MAPK) activation is a defining feature of ATRA-induced HL-60 differentiation.
ATRA drives sustained activation of the Raf/MEK/ERK pathway, leading to arrest and differentiation \cite{Yen1998}.
Betraying a feedback loop, MEK inhibition results in the loss of ERK and Raf phosphorylation, and the failure to arrest and differentiate in response to ATRA \cite{Hong2001}.
Retinoic acid (and its metabolites) are ligands for the hormone activated nuclear transcription factors retinoic acid receptor (RAR) and retinoid X receptor (RXR) \cite{Mangelsdorf1990}.
RAR/RXR activation is necessary for ATRA-induced Raf phosphorylation  \cite{Hong2001},
and the formation of an ATRA-inducible signalsome complex at the membrane, which drives MAPK activation.
While the makeup of the signalsome complex is not yet known,
we do know that it is composed of Src family kinases Fgr and Lyn, PI3K, c-Cbl, Slp76, and KSR, plus transcription factors
AhR and IRF1 \cite{Congleton:2012fk,Shen:2011aa,Shen:2009aa,Yen2006,Marchisio:1998aa}.
Signalsome activity is driven by ATRA-induced expression of CD38 and putatively the heterotrimeric Gq protein-coupled receptor BLR1 \cite{Congleton2011,WANG2004}.
BLR1 (also known as CXCR5), identified as an early ATRA (or D3)-inducible gene using differential display \cite{YEN1990},
is necessary for MAPK activation and differentiation \cite{WANG2004}, and drives signalsome activity.
Studies of the BLR1 promoter identified a non-canonical RARE site consisting of a 17 bp GT box approximately
1 kb upstream of the transcriptional start that conferred ATRA responsiveness \cite{WANG2004}.
Members of the BLR1 transcriptional activator complex, e.g. NFATc3 and CREB,
are phosphorylated by ERK, JNK or p38 MAPK family members suggesting positive feedback between the signalsome and MAPK activation \cite{Yang2002}.
BLR1 overexpression enhanced Raf phosphorylation and accelerated terminal differentiation, while Raf inhibition reduced BLR1 expression and ATRA-induced terminal differentiation \cite{Wang2008}.
In particular, Raf phosphorylation of the NFATc3 transcription factors at the BLR1 promoter enables transcriptional activation at the RARE by ATRA bound to RAR/RXR \cite{Geil:2014aa}.
BLR1 knock-out cells failed to activate Raf or differentiate in the presence of ATRA \cite{Wang2008}.
Interestingly, both the knockdown or inhibition of Raf, also reduced BLR1 expression and functional differentiation \cite{Wang2008}.
Thus, the expression of signalsome components e.g., BLR1 was Raf dependent, while Raf activation depended upon the signalsome.
A previous computational study of ATRA-induced differentiation of HL-60 cells suggested that the BLR1-MAPK positive feedback circuit was sufficient
to explain ATRA-induced sustained MAPK activation, and the expression of a limited number of functional differentiation markers \cite{Tasseff2011}.
Model analysis also suggested that Raf was the most distinct of the MAPK proteins.
However, this previous study developed and analyzed a complex model,
thus leaving open the critical question of what is the minimal positive feedback circuit required to drive ATRA-induced differentiation.

@@

\section*{Results}
We constructed an effective model of ATRA-induced HL-60 differentiation which described signaling and gene expression events following the addition of ATRA (Fig. \ref{fig:network}).
The model connectivity was developed from literature and the studies presented here (Table \ref{tbl:TF-network-connectivity}).
We decomposed the ATRA program into three modules;
a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RAR/RXR (Trigger) signals (Fig. \ref{fig:network}A);
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals (Fig. \ref{fig:network}B); and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors (Fig. \ref{fig:network}C).
In particular, Trigger (a surrogate for the RAR$\alpha$/RXR transcriptional complex)
regulated the expression of the transcription factors CCATT/enhancer binding protein $\alpha$ (C/EBP$\alpha$), PU.1, and Egr-1.
In turn, these transcription factors, in combination with cRaf-pS621, regulated the expression of downstream phenotypic markers such as CD38, CD11b or p47Phox.
Each component of these modules was described by a mRNA and protein balance equation.
Additionally, the signal initiation module also described the abundance of activated species e.g., Trigger and cRaf-pS621 whose values were derived from
unactivated Trigger and cRaf protein levels.
Lastly, because the population of HL-60 cells was dividing, we also considered a dilution term in all balance equations.
The signal initiation module contained nine differential equations, while the signal integration and phenotype modules were collectively encoded by 54 differential equations.
Model parameters were taken from literature (Table \ref{tbl:model-parameters}), or estimated from experimental data using heuristic optimization (see materials and methods).

@@

The signal integration and phenotype modules described ATRA-induced gene expression in wild-type HL-60 cells (Fig. \ref{fig:model-grn-simulations}).
The signal initiation module produced two outputs, activated Trigger and cRaf-pS621 which drove the expression of ATRA-induced transcription factors,
which then in turn activated the phenotypic program.
We assembled the connectivity of the signal integration and phenotypic programs driven by Trigger and cRaf-pS621 from literature (Table \ref{tbl:TF-network-connectivity}).
We estimated the parameters for the signal initiation, and phenotype modules
from steady-state and dynamic measurements of transcription factor and phenotypic marker expression following the addition of ATRA \cite{Jensen:2013aa,Jensen:2014aa,Jensen:2015ab,Jensen:2015aa}.
However, the bulk of the model parameters were taken from literature \cite{Milo:2010aa} and were not estimated in this study (see materials and methods).
The model simulations captured the time dependent expression of CD38 and CD11b following the addition ATRA (Fig. \ref{fig:model-grn-simulations}A),
and the steady-state for signal integration and phenotypic markers (Fig. \ref{fig:model-grn-simulations}B).
Lastly, we used the \textit{predicted} values of the p21 and E2F protein abundance to estimate a blackbox model of ATRA-induced G0 arrest (Fig. \ref{fig:model-cell-cycle-simulations}).
The phenotype module predicted p21 expression significantly increased and E2F expression decreased, in response to ATRA exposure (Fig. \ref{fig:model-cell-cycle-simulations}A).
We then used the ratio of these values in a polynomial model to calculate the fraction of HL-60 cells in G0 arrest following the addition of ATRA (Fig. \ref{fig:model-cell-cycle-simulations}B).
The third-order polynomial model captured the trend in measured G0-arrest values as a function of time, and was robust to uncertainty in the measured data (Fig. \ref{fig:model-cell-cycle-simulations}B, gray).
Taken together, the output of the signal integration and phenotypic modules was consistent with time-series and steady-state measurements, thereby validating the assumed molecular connectivity.
Moreover, outputs from the phenotype module described the trend in ATRA-induced G0 cell cycle arrest.
Next, we explored which proteins and protein interactions in the signal integration module most influenced the system response.

@@

The Gfi-1 and PPAR$\gamma$ proteins were important regulators of ATRA-induced signal integration and phenotypic change (Fig. \ref{fig:model-response-ko}).
We conducted pairwise gene knockout simulations in the signal integration and phenotype modules to estimate which proteins
controlled the processing of the Trigger and cRaf-S621 signals.
The difference between the system state with and without the gene knockouts (encoded as a normalized state displacement matrix) was decomposed using Singular Value Decomposition (SVD).
A panel of ten parameter sets was sampled, and the average normalized displacement matrix was decomposed.
The first six modes (approximately 36\% of the total) described $>$95\% of the gene knockout variance,
with the most important components of these modes being the Gfi-1 and PPAR$\gamma$ proteins, and to a lesser extent PU.1, C/EBP$\alpha$ and and AP1 (Fig. \ref{fig:model-response-ko}A).
To better understand which protein-DNA connections were important, we simulated the pairwise deletion of interactions between these proteins and their respective regulatory targets.
Singular value decomposition of the normalized state displacement matrix assembled from the pairwise connection deletions, suggested the first six modes (approximately 26\% of the total) accounted for $>$90\% of the variance.
Globally, the most sensitive interactions controlled p21 and p47Phox expression, markers for cell-cycle arrest and reactive oxygen formation phenotypic axes activated following ATRA addition (Fig. \ref{fig:model-response-ko}B).
Analysis of the modes suggested the action of PPAR$\gamma$, Gfi-1 and C/EBP$\alpha$ were consistently important over multiple target genes.
The connection knockout analysis also revealed robustness in the network.
For example, no pair of deletions qualitatively changed the expression of regulators such as PU.1, Oct1, Oct4 or PPAR$\gamma$.
Thus, the expression of these species was robust to disturbance in the connectivity.
To better understand the combined influence of the PPAR$\gamma$ and Gfi-1 deletions, we computed the fold change in the protein levels in the
single (Gfi-1$^{-/-}$ or PPAR$\gamma$$^{-/-}$) and double (Gfi-1$^{-/-}$ and PPAR$\gamma$$^{-/-}$) mutants for the best fit parameter set (Fig. \ref{fig:model-response-ko-simulations}).
Deletion of Gfi-1 led to a 2-4 fold increase in EGR-1, CD11b and C/EBP$\alpha$ expression, and a $>$8 fold increase in PU.1 abundance (Fig. \ref{fig:model-response-ko-simulations},blue).
On the other hand, deletion of PPAR$\gamma$ led to $>$8 fold downregulation of CD38, p21, IRF1 and Oct1 (Fig. \ref{fig:model-response-ko-simulations},red).
Both knockouts slightly increased E2F expression, but neither influenced the expression of p47Phox.
The double mutant was qualitatively similar to the combined behavior of the two single mutant cases.
Taken together, Gfi-1 and PPAR$\gamma$ controlled the cell-cycle arrest and receptor signaling axes, with PPAR$\gamma$ regulating CD38, IRF1 and p21 expression
while Gfi-1 controlled CD11b expression. These simulations suggested deletion of PPAR$\gamma$ and Gfi-1 would not interfere with reactive oxygen formation,
but would limit the ability of HL-60 cells to arrest.
However, this analysis did not give insight into which components upstream of the signal initiation module were important.
Toward this question, we explored the composition and regulation of the signalsome complex by experimentally interrogating a panel of possible Raf interaction partners.

@@

The full composition of the signalsome, and the kinase therein ultimately responsible for mediating ATRA-induced Raf activation is currently unknown.
To explore this question, we conducted immunoprecipitation and subsequent Western blotting to identify physical interactions between Raf and 19 putative interaction partners.
A panel of 19 possible Raf interaction partners (kinases, GTPases, scaffolding proteins etc)
was constructed based upon known signaling pathways.
We did not consider the most likely binding partner,
the small GTPase RAS, as previous studies have ruled it out in MAPK activation in HL-60 cells \cite{Wang2008,Katagiri1994}.
Total Raf was used as a bait protein for the immunoprecipitation studies.
Interrogation of the Raf interactome suggested Vav1 was involved with ATRA-induced initiation of MAPK activity (Fig. \ref{fig:ip-data}).
Western blot analysis using total Raf and Raf-pS621 specific antibodies confirmed the
presence of the bait protein, total and phosphorylated forms, in the immunoprecipitate (Fig. \ref{fig:ip-data}A).
Of the 19 proteins sampled, Vav1, Src, CK2, Akt, and 14-3-3 co-precipitated with Raf, suggesting their co-existence in a complex was possible.
However, only the associations between Raf and Vav1, and Raf and Src were ATRA-inducible (Fig. \ref{fig:ip-data}).
Furthermore, the Vav1 and Src associations were correlated with Raf-pS621 abundance in the precipitate.
Other proteins e.g., CK2, Akt and 14-3-3, generally bound Raf regardless of phosphorylation status or ATRA treatment.
The remaining 14 proteins were expressed in whole cell lysate (Fig. \ref{fig:ip-data}B),
but were not detectable in the immuno-precipitate with Raf IP; consistent with the potential importance of the Raf-Vav interaction for signaling, it paralleled Raf phosphorylation at S621, a putative telltale of the activated kinase.
Treatment with the Raf kinase inhibitor GW5074 following ATRA exposure reduced the association of both Vav1 with Raf and Src with Raf (Fig. \ref{fig:ip-data}),
although the signal intensity for Src was notably weak.
However, GW5074 did not influence the association of CK2 or 14-3-3 with Raf, further demonstrating their independence from Raf phosphorylation.
Interestingly, the Raf-Akt interaction qualitatively increased following treatment with GW5074;
however, it remained unaffected by treatment with ATRA.
Src family kinases are known to be important in myeloid differentiation \cite{Miranda2007} and their role in HL-60 differentiation has been investigated elsewhere \cite{Congleton:2012fk}.
Given the existing work and variable reproducibility in the context of the Raf immunoprecipitate,
we did not investigate the role of Src further in this study.
Taken together, the immunoprecipitation and GW5074 results implicated Vav1 association to be correlated with Raf activation following ATRA-treatment.
Previous studies demonstrated that a Vav1-Slp76-Cbl-CD38 complex plays an important role in ATRA-induced MAPK activation and differentiation of HL-60 cells \cite{Shen:2009aa}.
Here we did not observe direct interaction of Raf with Cbl or Slp76;
however, this interaction could could be involved upstream.
Next, we considered the effect of the Raf kinase inhibitor GW5074 on functional markers of ATRA-induced growth arrest and differentiation.

@@

% The presence of Vav1 in Raf/Grb2 complexes has been shown to correlate with increased Raf activity in mast cells \cite{Song1996}.
% Furthermore, studies on Vav1 knockout mice demonstrated that the loss of Vav1 resulted in deficiencies
% of ERK signaling for both T-cells as well as neutrophils \cite{Costello1999,Graham2007}.
% While its function in the signalsome is unclear, Vav1 has been shown to associate with a Cbl-Slp76-CD38 complex in an ATRA-dependent manner;
% furthermore, transfection of HL-60 cells with Cbl mutants that fail to bind CD38, yet still bind Slp76 and Vav1, prevented
% ATRA-induced MAPK activation \cite{Shen:2009aa}.
% Thus, interaction of Cbl-Slp76-Vav1 and CD38 appears to be required for transmission of the ATRA signal by the signalsome.

\section*{Discussion}
In this study, we presented an effective model of ATRA-inducible differentiation of HL-60 cells.
The model consisted of three modules: a signal initiation module that sensed and transformed the ATRA signal into activated cRaf-pS621 and the ATRA-RAR/RXR (Trigger) signals;
a signal integration module that controlled the expression of upstream transcription factors given cRaf-pS621 and activated Trigger signals; and
a phenotype module which encoded the expression of functional differentiation markers from the ATRA-inducible transcription factors.
The model described the transcription and translation of genes in each module, and signaling events in each module in a growing population of HL-60 cells.
Model parameters were taken from literature, however, unknown coefficients
that appear in the promoter logic models were estimated from protein measurements in HL-60 cells following ATRA exposure.
Despite its simplicity, the effective model captured key features of the ATRA induced differentiation such as
sustained MAPK activation, and bistability with respect to ATRA exposure.
The model also described the expression of upstream transcription factors which regulated the expression of differentiation markers.
Lastly, analysis of the response of the model to perturbations identified Gfi-1 and PPAR$\gamma$ as master regulators of ATRA-induced differentiation.
We also reported an ATRA-inducible component of the signalsome, Vav1.
Vav1 is a guanine nucleotide exchange factor for Rho family GTPases that activate pathways leading to actin cytoskeletal rearrangements and transcriptional alterations \cite{Hornstein:2004aa}.
The Vav1/Raf association correlated with Raf activity, was ATRA-inducible and decreased after treatment with the Raf inhibitor GW5074.

Naturally occurring cell fate decisions often incorporate reinforcing feedback and bistability \cite{Ferrell:2002aa,Laslo2006}.
One of the most well studied cell fate circuits is the Mos mitogen-activated protein kinase cascade in \textit{Xenopus} oocytes.
This cascade is activated when oocytes are induced by the steroid hormone progesterone \cite{Xiong:2003aa}.
The MEK-dependent activation of p42 MAPK stimulates the accumulation of the Mos oncoprotein, which in turn activates MEK, thereby closing the feedback loop.
This is similar to the signal initiation module presented here; ATRA drives signalsome formation, which activates MAPK, which in turn leads to more signalsome activation.
Thus, while HL-60 and \textit{Xenopus} oocytes are vastly different biological models, their cell fate programs share a similar architectural feature.
Reinforcing feedback and bistability has also been implicated in hematopoietic cell fate determination.
Laslo et al showed in nonmalignant myelomonocytic cells that the counter antagonistic repressors, Gfi-1 and Egr-1/2 (whose expression is tuned by PU.1 and C/EBP$\alpha$), encode a bistable switch that
results in a macrophage, neutrophil or a mixed lineage population depending upon PU.1 and C/EBP$\alpha$ expression \cite{Laslo2006}.
The current model contained the Gfi-1 and Egr-1/2 agonistic switch; however, its significance was unclear for HL-60 cells.
The expression of Gfi-1, Egr-1/2, C/EBP$\alpha$ and PU.1 was not consistent with the canonical lineage pattern expected from literature.
For example, Egr-1/2 expression (associated with a macrophage lineage) increased, while Gfi-1 expression (associated with a neutrophil lineage)
was unchanged following ATRA exposure.
Thus, HL-60 cells, which are a less mature cancer cell line, exhibited a non-canonical expression pattern.
Other unrelated cell fate decisions such as programmed cell death have also been suggested to be bistable \cite{Bagci:2006aa}.
Still more biochemical networks important to human health, for example the human coagulation or complement cascades, also feature strong positive feedback elements \cite{Luan:2007aa}.
Thus, while reinforcing feedback is often undesirable in human engineered systems, it is at the core of a diverse variety of
cell fate programs and other networks important to human health.
